---
title: "R Data Testing Examples"
output:
  html_document:
    df_print: paged
---

This R Markdown Notebook is a tutorial for how to do Data Testing, or Data
Validation, in R with examples using the `validate`
[package](https://cran.r-project.org/web/packages/validate/vignettes/cookbook.html).
Some of you may already be familiar with this package from the RSC templates
that we use.

The general idea of this package is to create a set of rules or expectations
about your data set and then check that the data set meets these rules or
expectations.

```{r setup}
library(validate)
```

## Variable Check Examples

In the examples below, we will check that certain constraints or properties
are met on a field-by-field or column-by-column basis using the `cars` data set.

```{r cars}
data(cars)
head(cars, 10)
```

The 7 rules we will set and check for this data set are to make sure:

- both the `speed` and `dist` columns are numeric
- both the `speed` and `dist` columns do NOT have negative values
- the ratio of `speed/dist` is between 0 and 1.5 (inclusive)
- both the `speed` and `dist` columns do not have missing or `NA` values

All of the records in this data set meet these constraints except for two
records that failed the `speed/dist` ratio rule.

```{r variable_check}
rules <- validator(
  is.numeric(speed),
  is.numeric(dist),
  speed >= 0,
  dist >= 0,
  in_range(speed / dist, 0, 1.5),
  !is.na(speed),
  !is.na(dist)
)
out <- confront(cars, rules)
summary(out)
```

The `plot` function displays a visual summary of which rules failed.

```{r variable_check_plot}
plot(out)
```

The `violating` function presents records that violate the record-level rules.
In this example, that means all rules except for the data type rules.
We can use this function to see which two records failed the `speed/dist` ratio
rule.

```{r variable_check_violating}
violating(cars, out[3:7])
```


## Uniqueness and Availability Examples

In the examples below, we will check column uniqueness and data availability
using the `samplonomy` data set.

```{r samplonomy}
data(samplonomy)
head(samplonomy, 10)
```


### Uniqueness 

First, we will check that each record has a unique combination of `region`,
`period`, and `measure`. Essentially we are defining what combination
of columns are the primary key (PK) in this data set and checking that the
data set does not violate a PK constraint.

When checking uniqueness, it is also recommended to check that the columns
included in the uniqueness rule do not contain missing or `NA` values as well
since the concept of a primary key traditionally does not allow for missing or
null values. Not checking for missing values in this columns could allow for
some edge cases to pass the uniqueness data validation tests that really should
be failing.

In general, primary key constraints should be defined in the SAM.
If they are defined at SAM definition then there is no possible way to have a
PK constraint violated when pulling that data into an R script and so these
types of tests should not be necessary.
However, there may be exceptions to that rule. For example, if the query to pull
data into the R script uses a join to applies some advanced filtering logic.
Please consider where you apply these types of data quality checks.
Ideally, they should be placed as far upstream as it makes sense to do so.

```{r uniqueness}
rule <- validator(
  !is.na(region),
  !is.na(period),
  !is.na(measure),
  is_unique(region, period, measure)
)
out <- confront(samplonomy, rule)
summary(out)
```

The `violating` function presents the records that violate the uniqueness rule
defined above.

```{r uniqueness_violating}
violating(samplonomy, out)
```


### Availability 

Next, we will test using the `contains_at_least` and `contains_exactly`
functions:

- that each `region` and `measure` in this data set have a row representing
every year from 2014 to 2019
- for each `measure` we have a row representing every year-quarter from 2014 to
2019 for the Agria and Induston regions only

29 records fail this first data availability constraint and 300 records fail
this second data availability constraint.

```{r availability}

# keyset for every year-quarter per measure for two regions
years <- as.character(2014:2019)
quarters <- paste0("Q", 1:4)
period_region_ai_keyset <- expand.grid(
  region = c("Agria", "Induston"),
  period = sapply(years, paste0, quarters)
)

rule <- validator(
  contains_at_least(
    keys = data.frame(period = as.character(2014:2019)),
    by = list(region, measure)
  ),
  contains_at_least(keys = minimal_keys, by = measure)
)

out <- confront(samplonomy,
                rule,
                ref = list(minimal_keys = period_region_ai_keyset))
summary(out)
```
